On the Synchronization Power of the OpenFlow Data Plane

### 2015-05-22 inband-sync.tex
completed first version

\newcommand{\addr}{\emph{addr}\xspace}
\newcommand{\pid}{\emph{pid}\xspace}
\newcommand{\add}{\textbf{add}\xspace}
\newcommand{\dele}{\textbf{delete}\xspace}

\newcommand{\FlowMod}{\emph{FlowMod}\xspace}
\newcommand{\match}{\emph{match}\xspace}
\newcommand{\action}{\emph{action}\xspace}
\newcommand{\flags}{\emph{flags}\xspace}
\newcommand{\checko}{\texttt{check\_overlap}\xspace}
\newcommand{\ufunc}{U} %apply\_update
\newcommand{\stefan}[1]{\textit{\textcolor{red}{[stefan]: #1}}} % Comments
\newcommand{\liron}[1]{\textit{\textcolor{mygreen}{[liron]: #1}}} % Comments
\newcommand{\petr}[1]{\textit{\textcolor{blue}{[petr]: #1}}} % Comments

@@

\algblockdefx[Bundle]{startTxn}{endTxn} %
[0]{\textbf{bundle}\{} %
[1][res]{ \} $\rightarrow #1$}

@@

We then show that, using our abstraction, the controllers can 
solve several important synchronization problems. 
We describe an implementation of the fundamental \emph{compare-and-swap} primitive, 
which in turn can be used to solve consensus and
implement a generic replicated service.
We then implement a simple CAS-based concurrent policy-update mechanism. 
Finally, we use our synchronization abstraction to provide
the missing link for the read-modify-write object
postulated in~\cite{cpc}).

@@

\begin{algorithm}[h]
    \caption{$\textit{\compare}(\addr,k)$}
    \label{alg:compare}
    \begin{algorithmic}[1]
    %\Require  id $i$.
%    \Ensure
    		\State $value \gets \addr$
    		\State $mask \gets  \texttt{1$\ldots$1}.\texttt{1$\ldots$1}$
    		\State $match \gets (value,mask)$
    		%\State $cookie \gets x\concat self$
    		\State $action \gets k$
    		\State $flag \gets \texttt{OFPFF\_\checko}$
    		\State $cmd\gets \emph{FlowMod}(match, op = \add, flag, action) $
			\Return $cmd$
    \end{algorithmic}
\end{algorithm}

@@

As a second example, let us consider a simple policy update protocol that
allows a set of  controllers to concurrently update switch  \emph{policies}, i.e., sets of
effective flow tables rules, under the
condition that these rules can be composed with the currently installed
policy.
Let us assume that the composition predicate is encoded in a user-specific
$\textit{\ufunc}()$ function~\cite{cpc}.
Algorithm~\ref{alg:simple-update} uses the bundle feature to update the switch configuration, where the configuration contains both the policy and an abstracted memory address that holds the \emph{policy identifier (pid)} .
In the algorithm, the controller first reads the currently installed
policy and its id \emph{pid}, executes the local update function ($\textit{\ufunc}$) which results with update commands and then tries to apply them atomically with updating the $pid$. If, meanwhile, a new
policy (with a different identifier) has been installed, the update
fails and takes no effect.
% \liron{the term composition doesn't make sense now with the new user update function... }

\liron{we can explain how a normal controller code can be instrumented to use our paradigm}

We use increasing numbers as policies
identifiers and use them in order to make sure that the new policy, with id
$pid+1$, replaces the (what considered as) current policy, with id
$pid$. Considering the policies identifiers, we send the policy update
commands as part of a bundle that also tries to perform
$CAS(pid,pid+1)$, thereby insuring that the commands are
performed iff the current policy is still there and preventing other
controllers to replace the new policy with another without observing it.

@@

\begin{algorithm}[h]
    \caption{Policy update with only CAS}
    \label{alg:simple-update}
    \begin{algorithmic}[1]
    \Require unsafe policy update function $\ufunc$, a predefined switch (memory) address for policy id: $\pid_{\addr}$
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
 			\State $\pid,policy\gets$ \textbf{read-config} %switch configuration
 			\State $update\_cmds\gets \ufunc(policy)$
 			\startTxn
	 			\State $CAS(\pid_{\addr},\pid,\pid+1)$,
	 			\State $update\_cmds$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
     		%\State $res \gets $ commit transaction
     	\Until{$res\neq \bot$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return

@@

Algorithm~\ref{alg:simple-update} allows policy identifiers to grow
arbitrarily. In fact, as was observed in~\cite{cpc}, the number of currently
used policies can be made proportional to the number of concurrent
controllers.

\hide{
It takes a set of rules $U$ (not yet a policy)and proceeds as follows: first, we seek to
 obtain a unique \emph{id}. FIXME: to be continued...
 \textbf{LS: I am not sure if I need to tell every step of the story or maybe it best to explain the main dif from the previous one similarly to what I just wrote next}.
}

Algorithm~\ref{alg:update} uses bounded number of policy ids by claiming the current policy id preventing someone of using it with different policy in the future (during the current update) and in the bundle used to update the policy it also checks that the new policy id that it about to use (chosen from a finite set of identifiers) wasn't claimed by someone else.

\hide{
We compute our new suggested policy by applying the update requests on top of current policy, supporting any kind of requests and policies. Then we make a transaction (using the bundle feature) to atomically check that our policy id is not blocked by anyone else, to change the current policy id to ours (an action that would fail if the current policy id is no longer what we are counting of) and to actually configure our new policy.

@@

\begin{algorithm}[h]
    \caption{Advanced policy update}
    \label{alg:update}
    \begin{algorithmic}[1]
        \Require unsafe policy update function $\ufunc$, a predefined switch (memory) address for policy id: $\pid_{\addr}$, a predefined id space $C$.
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
		 	\State $\pid,claims,policy\gets$ \textbf{read-config} %switch configuration
 			\State $claim(\pid)$
 			\State $\pid 2,claims,policy\gets$ \textbf{read-config} %switch configuration
 			\If {$\pid\neq \pid 2$}
	 			\State $unclaim(\pid)$
 				\State {\bf continue} (\textbf{restart} loop)
 			\EndIf
 			\State $my\_id\gets$ choose a number from $C\setminus claims$
 			\State $update\_cmds\gets \ufunc(policy)$
 			\startTxn
 				\State $\claimcheck(my\_id)$,
	 			\State $CAS(\pid_{\addr}, \pid,my\_id)$,
	 			\State $update\_cmds$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
	 		\State $unclaim(\pid)$
     		%\State $res \gets $ commit transaction
     	\Until{$res\neq\bot$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return

@@

Note that in the case of Algorithm \ref{alg:update}, switch configuration is used to store $pid$ and resource claims (both implemented by flow entries) in addition to the policy.

@@

\petr{this is the first time we talk concretely about "policies" ("forwarding policies"?), but they were mentioned a few times before. Shouldn't we have a short "SDN primer" in Sec 3?}

\petr{did not get what a "composition predicate" is}

\petr{why apply-update is "local" ("local update function")?}

\petr{the 1st paragraph is hard to digest}

\petr{Sec 5.2 relies too heavily on [3]. Do we expect the reader unfamiliar with [3] to get it?}

@@

Our work also contributes to the ongoing discussion of what can be implemented
in-band in OpenFlow~\cite{compute,reclaim}, by assuming the relevant perspective of distributed
synchronization.

@@

\section{Conclusion}\label{sec:conclusion}

@@

>>>>

\newcommand{\addr}{\textit{addr}\xspace}
\newcommand{\paddr}{\textit{paddr}\xspace}
\newcommand{\pid}{\textit{pid}\xspace}
\newcommand{\add}{\textbf{add}\xspace}
\newcommand{\dele}{\textbf{delete}\xspace}

\newcommand{\FlowMod}{\emph{FlowMod}\xspace}
\newcommand{\match}{\emph{match}\xspace}
\newcommand{\action}{\emph{action}\xspace}
\newcommand{\flags}{\emph{flags}\xspace}
\newcommand{\checko}{\texttt{check\_overlap}\xspace}
\newcommand{\ufunc}{update} %apply\_update
\newcommand{\stefan}[1]{\textit{\textcolor{red}{[stefan]: #1}}} % Comments
\newcommand{\liron}[1]{\textit{\textcolor{mygreen}{[liron]: #1}}} % Comments
\newcommand{\petr}[1]{\textit{\textcolor{blue}{[petr]: #1}}} % Comments

@@

\algblockdefx[Bundle]{startTxn}{endTxn} %
[0]{\textbf{\execatomic}\{} %
[1][res]{ \} $\rightarrow #1$}

@@

We then show that, using our abstraction, the controllers can 
solve several important synchronization problems. 
We describe an implementation of the fundamental \emph{compare-and-set} primitive, 
which in turn can be used to solve consensus and
implement a generic replicated service.
We then implement a simple CAS-based concurrent policy-update mechanism. 
Finally, we use our synchronization abstraction to provide
the missing link for the read-modify-write object
postulated in~\cite{cpc}).

@@

\begin{algorithm}[h]
    \caption{$\textit{\compare}(\addr,k)$}
    \label{alg:compare}
    \begin{algorithmic}[1]
    %\Require  id $i$.
%    \Ensure
    		\State $value \gets \addr$
    		\State $mask \gets  \texttt{1$\ldots$1}.\texttt{1$\ldots$1}$
    		\State $match \gets (value,mask)$
    		%\State $cookie \gets x\concat self$
    		\State $action \gets k$
    		\State $flag \gets \texttt{ofpff\_\checko}$
    		\State $cmd\gets \emph{FlowMod}(match, op = \add, flag, action) $
			\Return $cmd$
    \end{algorithmic}
\end{algorithm}

@@

As a second example, let us consider a simple policy update protocol that
allows a set of  controllers to concurrently update switch policies, i.e., sets of
%[[PK policies have been defined in Background]]
effective flow entries, under the
condition that these rules can be \emph{composed} with the currently installed
policy.

Let us assume that the composition predicate is encoded in a user-specific
$\textsc{\ufunc}()$ function~\cite{cpc}.
Algorithm~\ref{alg:simple-update} uses the bundle feature to update
the switch configuration, where the configuration contains both the
policy and a meta-configuration memory address $\paddr$ that holds the \emph{policy identifier (pid)} .
In the algorithm, the controller first reads the currently installed
policy together with its {\pid},  applies the
$\textsc{\ufunc}()$ function to the current policy which results in a
set of update FlowMod commands and then tries to apply them 
atomically together with a CAS operation on $\paddr$ replacing $\pid$
with $\pid+1$. 
The latter ensures that if, concurrently, a new policy (with a different identifier) has been installed, the update
fails and takes no effect.
% \liron{the term composition doesn't make sense now with the new user update function... }

\liron{we can explain how a normal controller code can be instrumented to use our paradigm}
\petr{did not get the comment}

%[[PK clear from the algorithm?]]
% and use them in order to make sure that the new policy, with id
%$pid+1$, replaces the (what considered as) current policy, with id
%$pid$. 
%Considering the policies identifiers, we send the policy update
%commands as part of a bundle that also tries to perform
%$CAS(pid,pid+1)$, thereby insuring that the commands are
%performed iff the current policy is still there and preventing other
%controllers to replace the new policy with another without observing it.

\petr{what does it mean for an update function to be ``unsafe''?}

@@

\begin{algorithm}[h]
    \caption{Policy update with only CAS}
    \label{alg:simple-update}
    \begin{algorithmic}[1]
    \Require unsafe policy update function $\ufunc$, policy id $\addr$
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
 			\State $\pid,policy\gets$ \textbf{read-config} %switch configuration
 			\State $update\_cmds\gets \ufunc(policy)$
 			\startTxn
	 			\State $CAS(\pid_{\addr},\pid,\pid+1)$,
	 			\State $update\_cmds$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
     		%\State $res \gets $ commit transaction
     	\Until{$res=\ack$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return

@@

Note that in Algorithm~\ref{alg:simple-update}, policies identifiers
grow without bound.
However, as was observed in~\cite{cpc}, the number of currently
used identifiers can be made proportional to the number of concurrent
controllers.
%
\hide{
It takes a set of rules $U$ (not yet a policy)and proceeds as follows: first, we seek to
 obtain a unique \emph{id}. FIXME: to be continued...
 \textbf{LS: I am not sure if I need to tell every step of the story or maybe it best to explain the main dif from the previous one similarly to what I just wrote next}.
}
%

Algorithm~\ref{alg:update} uses our id-claimer abstraction 
to bound the number of policy identifiers.
Claiming a policy id prevents another controller from using it with a
different policy.
A controller claims the current policy identifier, chooses a new id
from a finite set of identifiers,   
and then checks that the new policy id is not claimed by another controller.
Note that in Algorithm \ref{alg:update}, the meta-configuration is
used to store $pid$ and id-claimer.
\petr{The description is not very clear.}

\hide{
We compute our new suggested policy by applying the update requests on
top of current policy, supporting any kind of requests and
policies. Then we make a transaction (using the bundle feature) to
atomically check that our policy id is not blocked by another controller, to change the current policy id to ours (an action that would fail if the current policy id is no longer what we are counting of) and to actually configure our new policy.

@@

\begin{algorithm}[h]
    \caption{Advanced policy update}
    \label{alg:update}
    \begin{algorithmic}[1]
        \Require unsafe policy update function $\ufunc$, a predefined switch (memory) address for policy id: $\pid_{\addr}$, a predefined id space $C$.
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
		 	\State $\pid,claims,policy\gets$ \textbf{read-config} %switch configuration
 			\State $\textbf{execute}\{claim(\pid)\}$
 			\State $\pid 2,claims,policy\gets$ \textbf{read-config} %switch configuration
 			\If {$\pid\neq \pid 2$}
	 			\State $\textbf{execute} \{unclaim(\pid)\}$
 				\State {\bf continue} (\textbf{restart} loop)
 			\EndIf
 			\State $my\_id\gets$ choose a number from $C\setminus claims$
 			\State $update\_cmds\gets \ufunc(policy)$
 			\startTxn
 				\State $\claimcheck(my\_id)$,
	 			\State $CAS(\pid_{\addr}, \pid,my\_id)$,
	 			\State $update\_cmds$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
	 		\State $\textbf{execute} \{unclaim(\pid)\}$
     		%\State $res \gets $ commit transaction
     	\Until{$res=\ack$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return

@@

@@

@@

Our work also contributes to the ongoing discussion of what can be implemented
in-band in OpenFlow~\cite{compute,reclaim}, by focusing on the
relevant problem of synchronization.

@@

\section{Concluding Remarks}\label{sec:conclusion}

@@

Even though we believe that we propose efficient and consistent
transactional framework for in-band synchronization, we do not claim
that we offer an exhaustive set of synchronization primitives. Indeed,
one may explore standard OpenFlow features to enable more powerful
primitives. For instance, \emph{cookies}~\cite{openflow} can be used for modifying
flow tables, based on more sophisticated conditions than the default
``exact match''. 
Also, the standard pipelining feature allows the switches to maintain multiple flow
tables which can be traversed in arbitrary patterns, defining the sets
of corresponding actions. 
Exploring the power of such features is left to future work.
